\chapter*{Contributors}
\addcontentsline{toc}{chapter}{Contributors}

\subsection*{Sean Collins}
Sean Collins is a physical therapist with a doctor of science in ergonomics and epidemiology from the UMass Lowell. He was professor of physical therapy and biomedical engineering at UMass Lowell for 18 years prior to relocating to Plymouth State University (PSU) as the founding director of the Doctor of Physical Therapy (DPT) program where is currently a professor. This project is possible because of the PSU DPT curriculum, and the generous sabbatical support from the PSU administration and DPT faculty colleagues. He is the primary editor for the project and author of several chapters.

\subsection*{Bog Sniezek}
Bog Sniezek is a systems scientist (more to come). He is the supporting editor and founder of the systems science approach being utilized (Unified Systems Theory).

\subsection*{Tom Sniezek}
Tom Sniezek is a physical therapist, graduate of the DPT program at UMass Lowell and currently practices physical therapy in Ohio. (More on role to come)

\subsection*{Nathaniel Mailloux}
Nathaniel Mailloux is a physical therapist, graduate of the DPT program at Plymouth State University and currently practices physical therapy in New York. He is the primary author for the Integrative Exercise on Fick's Equation.

\subsection*{Kyle Coffey}
Bio coming soon. He is the primary author for the Integrative Exercise on Exercise Prescription; and co-author for the chapters in the section on Muscle Integrity.

\printbibliography[heading=subbibintoc]